---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

Author  : Akash Chauhan  
Date    : 19th April 2015


## Loading and preprocessing the data

The current directory is set to the forked repo directory.
Unzip the activity.zip file and then read the csv file using the following command - 

```{r}
activity <- read.csv("activity.csv")
```

A quick view of the loaded dataset :

```{r}
head(activity)
```


## What is mean total number of steps taken per day?

Find out the total no. of steps per day using the following code -

```{r}
act_sum <- aggregate(activity$steps, by=list(activity$date), FUN="sum")
names(act_sum) <- c("Date", "total_steps")
head(act_sum)
```

plot a histogram for the total no. of steps:

```{r}
hist(act_sum$total_steps,col="lightblue", main="Distribution of total steps per day",xlab="Total steps per day")
```

Mean and median of total steps per day:

```{r}
mean(act_sum$total_steps, na.rm=TRUE)
median(act_sum$total_steps, na.rm=TRUE)
```


## What is the average daily activity pattern?

calculate the average no. of steps by time interval across all dates:

```{r}
stepbyint <- aggregate(activity$steps, by=list(activity$interval), FUN="mean", na.rm=TRUE)
names(stepbyint) <- c("Interval", "avg_steps")
head(stepbyint)
plot(stepbyint$Interval, stepbyint$avg_steps, type="l", main="Plot of average steps by interval", xlab="Interval", ylab="average steps")
```

finding out the maximum steps interval:

```{r}
stepbyint[stepbyint$avg_steps == max(stepbyint$avg_steps),]$Interval
```

## Imputing missing values

Find out the no. of rows with NA values

```{r}
nrow(activity[is.na(activity),])
```


Replacing the missing NA values of steps with the median steps for that interval

```{r}
actwona <- activity[!is.na(activity$steps),]
stepbyint <- aggregate(actwona$steps, by=list(actwona$interval), FUN="median")
names(stepbyint) <- c("interval", "steps_new")
head(stepbyint)
```

creating a new dataset with the missing steps replaced with the new steps

```{r}
act1 <- merge(activity, stepbyint, by="interval")
act2 <- act1[!is.na(act1$steps),]
act3 <- act1[is.na(act1$steps),]
act2 <- act2[,c("interval","steps","date")]
act3 <- act3[,c("interval","steps_new","date")]
names(act3)[2] <- "steps"
act_new <- rbind(act2, act3)
act_new <- act_new[order(act_new$date, act_new$interval),]
head(act_new)
```


The following is a repeat of the previous steps to create a histogram:


```{r}
act_sum <- aggregate(act_new$steps, by=list(act_new$date), FUN="sum")
names(act_sum) <- c("Date", "total_steps")
head(act_sum)
hist(act_sum$total_steps,col="lightblue", main="Distribution of total steps per day",xlab="Total steps per day")
mean(act_sum$total_steps, na.rm=TRUE)
median(act_sum$total_steps, na.rm=TRUE)
```

The histogram and mean/median values clearly have changed from before.




## Are there differences in activity patterns between weekdays and weekends?


Add week day variable to the new dataset and create a panel plot for
either weekday or weekend.

```{r}
act_new$wkd <- weekdays(as.Date(act_new$date,format="%Y-%m-%d"))
act_new$wk <- ifelse(act_new$wkd == "Saturday", "Weekend", ifelse(act_new$wkd == "Sunday", "Weekend", "Weekday"))

par(mfrow=c(2,1))
act_new_wkd <- act_new[act_new$wk == "Weekday",]
act_new_wkend <- act_new[act_new$wk == "Weekend",]

stepbyint1 <- aggregate(act_new_wkd$steps, by=list(act_new_wkd$interval), FUN="mean", na.rm=TRUE)
names(stepbyint1) <- c("Interval", "avg_steps")
plot(stepbyint1$Interval, stepbyint1$avg_steps, type="l", main="Plot of average steps by interval for weekdays", xlab="Interval", ylab="average steps")


stepbyint2 <- aggregate(act_new_wkend$steps, by=list(act_new_wkend$interval), FUN="mean", na.rm=TRUE)
names(stepbyint2) <- c("Interval", "avg_steps")
plot(stepbyint2$Interval, stepbyint2$avg_steps, type="l", main="Plot of average steps by interval for weekend", xlab="Interval", ylab="average steps")
```


